---
title: "College_Football_Visualizations"
output: html_document
---
# Load necessary libraries and packages


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
if (!require("httr")) install.packages("httr")
if (!require("jsonlite")) install.packages("jsonlite")
if (!require("tidyverse")) install.packages("tidyverse")


# Load libraries
library(ggplot2)
library(dplyr)
library(forcats)
library(cowplot)
library(ggrepel) 
library(tidyr)
library(plotly)
library(httr)
library(jsonlite)
library(scales)
library(tidyverse)
library(stringr)
library(RColorBrewer)


```





# Connect to API 
```{r}

# Get API key from .Renviron
api_key <- Sys.getenv("CFB_API_KEY")

```





# Load data 

## Overall team talent
```{r}

# API call to team talent composite
res <- GET(
  url = "https://api.collegefootballdata.com/talent",
  add_headers(Authorization = paste("Bearer", api_key))
)

# Parse the JSON response
talent_data <- fromJSON(content(res, as = "text"), flatten = TRUE)

# Preview
head(talent_data)



```


## SP +  advanced ratings
```{r}

# API call to team talent composite
res <- GET(
  url = "https://api.collegefootballdata.com/ratings/sp",
  query = list(year = 2024),  # Optional: filter by year
  add_headers(Authorization = paste("Bearer", api_key))
)

# Parse the JSON response
sp_data <- fromJSON(content(res, as = "text"), flatten = TRUE)

# Preview
head(sp_data)


```



## Coaching history
```{r}

# API call to team talent composite
res <- GET(
  url = "https://api.collegefootballdata.com/coaches",
  query = list(year = 2024),  # Optional: filter by year,
  add_headers(Authorization = paste("Bearer", api_key))
)

# Parse the JSON response
coach_data <- fromJSON(content(res, as = "text"), flatten = TRUE)

# Preview
head(coach_data)


```



## Betting lines
```{r}

# API call to team talent composite
res <- GET(
  url = "https://api.collegefootballdata.com/lines",
  query = list(year = 2024),  # Optional: filter by year,
  add_headers(Authorization = paste("Bearer", api_key))
)

# Parse the JSON response
line_data <- fromJSON(content(res, as = "text"), flatten = TRUE)

# Preview
head(line_data)


```



## Team recruiting ranks
```{r}

# API call to team talent composite
res <- GET(
  url = "https://api.collegefootballdata.com/recruiting/teams",
  query = list(year = 2024),
  add_headers(Authorization = paste("Bearer", api_key))
)

# Parse the JSON response
recruit_data <- fromJSON(content(res, as = "text"), flatten = TRUE)

# Preview
head(recruit_data)


```


## Returning player production
```{r}

# API call to team talent composite
res <- GET(
  url = "https://api.collegefootballdata.com/player/returning",
  query = list(year = 2024),
  add_headers(Authorization = paste("Bearer", api_key))
)

# Parse the JSON response
exp_data <- fromJSON(content(res, as = "text"), flatten = TRUE)

# Preview
head(exp_data)


```


# A chart that shows the distribution of a single categorical variable

## Distribution of college football conferences for 2024 season

```{r}

# Inspect data
glimpse(sp_data)
glimpse(recruit_data)
glimpse(coach_data)
glimpse(talent_data)
glimpse(exp_data)

# Sort conferences by frequency
sp_df <- sp_data %>%
  filter(!is.na(conference)) %>%
  mutate(conference = as.factor(conference)) %>% 
  mutate(conference = fct_lump(conference, n=9)) %>%
  count(conference, sort = TRUE)


# Bar chart

# Create a palette
text_palette <- colorRampPalette(c("darkgray", "darkorange"))

sp_df <- sp_df %>%
  mutate(
    rank = row_number(),
    skip_color = rank > (n() - 2),
    alt_index = rank %% 2 == 1,
    text_color = ifelse(
  skip_color, 
  "transparent",  # hide labels for last 2
  ifelse(rank %in% c(1, 2, 3, 5, 7),  # always show for given ranks
         text_palette(n())[rank],
         "transparent")  # hide for everything else

    )
  )


p <- ggplot(data = sp_df, aes(x = fct_reorder(conference, n), y = n, fill = n)) + geom_col(show.legend = FALSE) +
  geom_text(aes(label = n, color = text_color),
            hjust = 1.1,
            size = 4,
            show.legend = FALSE) +
  scale_color_identity() +
  coord_flip() + scale_fill_gradient(low = "lightblue", high = "darkblue") + labs(
    title = "Number of Teams by College Football Conference",
    subtitle = "2024-25 Season",
    x = NULL,
    y = "Number of Teams",
    fill = "Count"
  ) + theme_minimal()
  
p
  
```


### Delete after interpreting the plot: The number of teams for the conference ranges from nearly 18 to 
    The last 2 teams to win the College Football National Championship are from the 10, Ohio state in 2025
    and Michigan in 2024. 



# A chart that shows the distribution of a single quantitative variable
## Distribution of offensive efficiency for College Football Conferences from the 2024-25 season

```{r}

# Filter out NA
big10_df <- exp_data %>%
  filter(!is.na(conference) )

# Inspect data Frame
glimpse(big10_df)


# Stacked bar chart of home_ownership  vs application_type

x_limits <- range(big10_df$percentPPA, na.rm = TRUE)
binwidth_percent <- (max(big10_df$percentPPA, na.rm = TRUE) - min(big10_df$percentPPA, na.rm = TRUE)) / 15
median_val <- median(big10_df$percentPPA, na.rm = TRUE)

hist_p <- ggplot(big10_df, aes(x =  percentPPA)) +
  geom_histogram(binwidth = binwidth_percent, center = 0,
                 color = "darkgray", fill = "lightblue") +
  stat_function(fun = function(x) {
    dnorm(x, mean = mean(big10_df$percentPPA, na.rm = TRUE),
              sd = sd(big10_df$percentPPA, na.rm = TRUE)) * nrow(big10_df) * binwidth_percent},  
    aes(color = "Normal"), size = 1.2) +
  geom_density(aes(y = ..density.. * nrow(big10_df) * binwidth_percent, color = "Empirical"), 
               adjust = 1.5, size = 1.2, fill = NA) +
  scale_color_manual(name = "Density:", 
                     values = c("Normal" = "red", "Empirical" = "blue")) +
  labs( title = "Distribution of Offensive Efficiency for College Football Teams",  subtitle = "2024-25 Season",
        x = "percentPPA",y = "Count" ) + coord_cartesian(xlim = x_limits) + 
  theme_minimal()


box_p <- ggplot(big10_df, aes(x = percentPPA, y = factor(1))) +
  geom_boxplot(fill = "lightblue", width = 0.4, outlier.color = "black") +
   geom_text(aes(x = median_val, y = 1, label = paste0("Median = ", round(median_val, 2))),
            vjust = -0.000001, size = 4, color = "black") +  
  coord_cartesian(xlim = x_limits) +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.margin = margin(t = -15)  # reduce white space between plots
  ) +
  labs(y = NULL, x = NULL)

final_p <- plot_grid(hist_p, box_p, 
                     ncol = 1, 
                     align = "v",       # align vertically
                     rel_heights = c(3, 0.5))

final_p
  
```

# A chart that shows the distribution of two categorical variables
## Distribution of team ranking and conferences in College Football for 2024-25 season




```{r}

library(dplyr)
library(stringr)
library(plotly)

# 👉 Step 1: Define top conferences explicitly
top_confs <- c(
  "SEC", "Big Ten", "ACC", "Pac-12", "Big 12",
  "American Athletic", "Mountain West", "Conference USA", 
  "FBS Independents", "Mid-American", "Sun Belt"
)

# 👉 Step 2: Clean and categorize data
df_clean <- sp_data %>%
  filter(!is.na(ranking), !is.na(conference)) %>%
  mutate(
    conference = str_squish(str_trim(conference)),
    
    # 👉 Standardize known names
    conference = case_when(
      str_detect(conference, regex("Mountain", ignore_case = TRUE)) ~ "Mountain West",
      str_detect(conference, regex("SEC", ignore_case = TRUE)) ~ "SEC",
      str_detect(conference, regex("Big Ten", ignore_case = TRUE)) ~ "Big Ten",
      str_detect(conference, regex("ACC", ignore_case = TRUE)) ~ "ACC",
      str_detect(conference, regex("Pac", ignore_case = TRUE)) ~ "Pac-12",
      str_detect(conference, regex("Big 12", ignore_case = TRUE)) ~ "Big 12",
      str_detect(conference, regex("American", ignore_case = TRUE)) ~ "American Athletic",
      str_detect(conference, regex("Conference USA", ignore_case = TRUE)) ~ "Conference USA",
      str_detect(conference, regex("Independent", ignore_case = TRUE)) ~ "FBS Independents",
      str_detect(conference, regex("Mid-American", ignore_case = TRUE)) ~ "Mid-American",
      str_detect(conference, regex("Sun Belt", ignore_case = TRUE)) ~ "Sun Belt",
      TRUE ~ "Non-Top Conference"   # ✅ label for all other conferences
    ),
    
    # 👉 Collapse into top or non-top conferences
    conference = ifelse(conference %in% top_confs, conference, "Non-Top Conference"),
    
    # 👉 Define ranking groups
    ranking_group = case_when(
      ranking >= 1 & ranking <= 5 ~ "1-5",
      ranking >= 6 & ranking <= 10 ~ "6-10",
      ranking >= 11 & ranking <= 15 ~ "11-15",
      ranking > 15 ~ "Outside Top 15"
    ),
    ranking_group = factor(ranking_group, levels = c("1-5", "6-10", "11-15", "Outside Top 15"))
  )

# 👉 Step 3: Summarize for plotting
df_plot <- df_clean %>%
  group_by(ranking_group, conference) %>%
  summarise(
    count = n(),
    teams = paste(team, collapse = "<br>"),
    .groups = "drop"
  ) %>%
  group_by(ranking_group) %>%
  mutate(
    percent = 100 * count / sum(count)
  ) %>%
  arrange(ranking_group, desc(percent)) %>%
  ungroup()

# 👉 Step 4: Distinct color map
color_map <- c(
  "SEC" = "#E69F00",
  "Big Ten" = "#56B4E9",
  "ACC" = "darkgreen",
  "Pac-12" = "purple",
  "Big 12" = "#D55E00",
  "American Athletic" = "lightpink",
  "Mountain West" = "#999999",
  "Conference USA" = "#66C2A5",
  "FBS Independents" = "#000000",
  "Mid-American" = "#A52A2A",
  "Sun Belt" = "#20B2AA",
  "Non-Top Conference" = "gray40"  # ✅ for grouped others
)

# 👉 Step 5: Build plotly stacked bar chart
p_plotly <- plot_ly()
ranking_levels <- levels(df_clean$ranking_group)
used_legends <- c()

for (grp in ranking_levels) {
  df_grp <- df_plot %>%
    filter(ranking_group == grp) %>%
    arrange(percent)  # ensures tallest stack appears on top
  
  for (i in 1:nrow(df_grp)) {
    row <- df_grp[i, ]
    show_legend <- !(row$conference %in% used_legends)
    used_legends <- c(used_legends, row$conference)
    
    p_plotly <- add_trace(
      p_plotly,
      x = row$ranking_group,
      y = row$percent,
      name = row$conference,
      type = 'bar',
      marker = list(color = color_map[[row$conference]]),
      text = paste0(
        "<b>Conference:</b> ", row$conference,
        "<br><b>Percent:</b> ", round(row$percent, 1), "%",
        "<br><b>Teams:</b><br>", row$teams
      ),
      hoverinfo = "text",
      textposition = "none",
      showlegend = show_legend,
      legendgroup = row$conference
    )
  }
}

# 👉 Step 6: Layout
p_plotly <- layout(
  p_plotly,
  barmode = "stack",
  title = list(
    text = "<b>Top 15 Ranking Distribution by Conference</b><br><span style='font-size:12pt;'>2024–2025 Season</span>",
    x = 0.5
  ),
  xaxis = list(title = "Ranking Group", categoryorder = "array", categoryarray = ranking_levels),
  yaxis = list(title = "Percent (%)"),
  legend = list(title = list(text = "<b>Conference</b>")),
  margin = list(t = 80)
)

p_plotly




```


Power Five (P5) Conferences:
These are officially recognized by the NCAA, media, and CFP (College Football Playoff) as the most competitive and financially dominant leagues:

SEC

Big Ten

ACC

Big 12

Pac-12 (Note: dissolved for 2024–25, only Oregon State & Washington State remain independent or temporarily aligned)


These conferences historically receive automatic access to top bowl games, recruit elite players, and dominate TV revenue.


 Group of Five (G5) Conferences – strongest among non-P5:
American Athletic Conference (AAC) – top G5 league, had multiple ranked teams in recent years (e.g., Tulane, SMU, UCF before they moved)

Mountain West – strong G5 league with consistent competitiveness (e.g., Boise State, Air Force)

Conference USA – less competitive, but still in FBS; included for completeness

FBS Independents – like Notre Dame, Army, and formerly BYU; some (Notre Dame) compete at Power levels

Sun Belt – increasingly strong (e.g., App State, Troy)

Mid-American (MAC) – generally weaker in national rankings






  

A chart that shows the distribution of a quantitative variable across categories of a
categorical variable

# I will compare offense.ranking across the categories of conference 
```{r}


sp_data
glimpse(sp_data)

# Filter and clean offense.ranking and conference
df_box <- sp_data %>%
  filter(!is.na(offense.ranking), !is.na(conference)) %>%
  mutate(conference = str_squish(str_trim(conference)))  # Clean up spaces

df_box <- df_box %>% mutate(conference = fct_lump(factor(conference), n = 5))


# Create boxplot
Pbox <- ggplot(data = df_box, aes(x = conference, y = offense.ranking, fill = conference)) +
  geom_boxplot() +
  labs(
    x = "",
    y = "Offensive Ranking",
    title = "Distribution of Offensive Rankings by Conference"
  ) +
  theme_minimal() +
  theme(
    axis.text.x  = element_text(face = "bold", size = 10, angle = 45, hjust = 1),
    axis.title.y = element_text(face = "bold", size = 14),
    plot.title   = element_text(hjust = 0.5, face = "bold")
  ) +
  scale_fill_viridis_d(option = "plasma") +
  scale_y_reverse(expand = expansion(mult = c(0, 0.1))) +  # Optional: lower rankings are better
  theme(legend.position = "none")

Pbox


```





A chart that shows the relationship between two quantitative variables
Scatter plot of defense.rating and ranking  

```{r}


# Filter and select relevant variables
df_def <- sp_data %>%
  filter(!is.na(defense.rating), !is.na(ranking)) %>%
  select(team, ranking, defense.rating, offense.ranking, conference)

# Calculate IQR-based outlier bounds
Q1_def <- quantile(df_def$defense.rating, 0.25)
Q3_def <- quantile(df_def$defense.rating, 0.75)
IQR_def <- Q3_def - Q1_def
lb_def <- Q1_def - 1.5 * IQR_def
ub_def <- Q3_def + 1.5 * IQR_def

Q1_rank <- quantile(df_def$ranking, 0.25)
Q3_rank <- quantile(df_def$ranking, 0.75)
IQR_rank <- Q3_rank - Q1_rank
lb_rank <- Q1_rank - 1.5 * IQR_rank
ub_rank <- Q3_rank + 1.5 * IQR_rank

# Label outliers
df_def <- df_def %>%
  mutate(outlier_status = ifelse(
    (defense.rating < lb_def | defense.rating > ub_def) |
    (ranking < lb_rank | ranking > ub_rank),
    "Outlier", "Normal"
  ))

# Create interactive plot
p_plotly <- plot_ly(
  data = df_def,
  x = ~ranking,
  y = ~defense.rating,
  type = 'scatter',
  mode = 'markers',
  color = ~outlier_status,
  colors = c("blue", "orange"),
  text = ~paste(
    "<b>Team:</b>", team,
    "<br><b>Conference:</b>", conference,
    "<br><b>Offense Ranking:</b>", offense.ranking,
    "<br><b>Defense Rating:</b>", round(defense.rating, 2),
    "<br><b>Ranking:</b>", ranking
  ),
  hoverinfo = 'text',
  marker = list(size = 10, opacity = 0.8, line = list(width = 1, color = '#333'))
) %>%
  layout(
    title = list(
      text = "<b>Defense Rating vs Ranking</b><br><span style='font-size:12pt;'>2024–2025 Season</span>",
      x = 0.5,
      xanchor = "center"
    ),
    margin = list(t = 80),
    xaxis = list(title = "Ranking"),
    yaxis = list(title = "Defense Rating"),
    legend = list(title = list(text = "<b>Outlier Status</b>"))
  )

p_plotly





```




Radar/Spider Chart to display the mean rankings and special teams rating by conference

```{r}

glimpse(sp_data)



# Prepare data
radar_df <- sp_data %>%
  filter(!is.na(conference), 
         !is.na(offense.ranking), 
         !is.na(defense.ranking), 
         !is.na(specialTeams.rating)) %>%
  group_by(conference) %>%
  summarise(
    Offense = mean(offense.ranking, na.rm = TRUE),
    Defense = mean(defense.ranking, na.rm = TRUE),
    SpecialTeams = mean(specialTeams.rating, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Normalize values to [0,1] for fair comparison
  mutate(
    Offense = scales::rescale(Offense, to = c(0, 1)),
    Defense = scales::rescale(Defense, to = c(0, 1)),
    SpecialTeams = scales::rescale(SpecialTeams, to = c(0, 1))
  ) %>%
  pivot_longer(cols = -conference, names_to = "Metric", values_to = "Value")

# Unique conferences
conference_levels <- unique(radar_df$conference)

# Generate a color palette with as many distinct colors as conferences
num_colors <- length(conference_levels)
colors <- RColorBrewer::brewer.pal(min(num_colors, 8), "Set2")
if (num_colors > 8) {
  # extend colors using distinct hue palette if needed
  colors <- colorspace::rainbow_hcl(num_colors)
}
names(colors) <- conference_levels

# Create plot
radar_plot <- plot_ly(type = 'scatterpolar', fill = 'toself')

for (conf in conference_levels) {
  radar_plot <- radar_plot %>%
    add_trace(
      r = radar_df$Value[radar_df$conference == conf],
      theta = radar_df$Metric[radar_df$conference == conf],
      name = conf,
      line = list(color = colors[conf]),
      fillcolor = colors[conf],
      opacity = 0.5
    )
}

radar_plot <- radar_plot %>%
  layout(
    title = list(
      text = "<b>Average Rankings & Special Teams Ratings by Conference</b><br><span style='font-size:12pt;'>2024–2025 Season</span>",
      x = 0.5,
      xanchor = "center"
    ),
    margin = list(t = 80),
    polar = list(
      radialaxis = list(
        visible = TRUE,
        range = c(0, 1),
        tickfont = list(size = 10)
      )
    ),
    legend = list(
      orientation = "h",
      x = 0.5,
      xanchor = "center",
      y = -0.2,
      font = list(size = 11)
    )
  )

radar_plot





```


Lower ranking values indicate better performance. You can see how balanced conferences are based on the 3 categories.